Translation of B Implementations to the LLVM: Specification

### 2014-04-25 DOCUMENTATION/DerivedTypes.tex
rewriting the initialisation

\subsection{Rules to be updated in first specification}

\paragraph{Becomes equal instructions} \textcolor{blue}{ must evaluate the
 target and the source of the assignment, and then copy the result of the 
latter in the former, using the LLVM instruction \llvm{store}. The evaluation 
of the target is defined with $\trad{}{lvalue}$, which yields the corresponding 
LLVM source code $l$ and the variable $p$ containing the assigned location. The 
evaluation of the source yields the corresponding code $r$, the variable $v$ 
holding the result, and its type $t$. }

\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef
  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Exp}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
\end{split}
\end{align*}

The rule lvalue is:
\begin{align*}
\lefteqn{\trad{n}{lvalue} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \IF \mbox{\B{n} is an ArrayItem} \THEN  \trad{n}{ArrayItem} \\
& \quad \IF \mbox{\B{n} is a StructItem} \THEN  \trad{n}{StructItem} \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

\textcolor{blue}{ The expressions have elements represented by pointers that need a special 
care. The attributions between these elements are made  directly by the memory 
address.  However, the expression resolution are made by parts, where the 
partial results are stored to temporary variables. Therefore, the expression 
resolution with pointers needs an additional instruction $LOAD$ for store the 
partial results. This translation is represented by $ValueItem$.}

\begin{align*}
\begin{split}
  \trad{n}{\B{ValueItem}} \isdef \\
   & \LET \ell = \lbl  \IN\\
   & \IF \B{n} \mbox{ is \B{StructItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{StructItem}} \\
   & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \\
   & \sep  \ell  \sep t'\\
\end{split}
\end{align*}

\section{Translation of Structures and Array}
\label{sec:StructArray}

Two approaches are possible to support arrays and structures. An approach is 
importing reusable libraries of AtelierB that are reported in the official 
document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. 
This approach is supported, but the generated code is more complex and it needs 
more intermediate steps to manipulate the data of imported components. 
Moreover, this approach allows dynamic allocation for arrays, but it is 
strongly not recommended for safety-critical systems. A second approach enjoys 
the direct correspondence between B0 native types and LLVM and this approach 
is presented here.

\subsection{Structure} \VGM{It is not implemented yet, because it requires generating an XML example.}
The structure declaration defines a list of variables to be placed under one 
name, allowing several variables to be accessed by one base name. This example  
initializes a structure, and gets and sets one value of xx'balance.

\begin{ccode}
xx := rec(id:1, balance:10);
xx'balance := xx'balance -1;
\end{ccode}
The lines 1 and 2 are represented in translation in LLVM code:

@@

The $StructItem$ rule gets an element value from the struct. The same rule is used on the left side of attribution as a $LValue$ and right side of attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure.

\begin{align*}
\begin{split}
  \trad{n}{\B{StructItem}} \isdef \\
   & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
   &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   &\quad  \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
   &\quad  \sep  \ell  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

The attribution from record structure to a structure variable, line 1 from last 
B code example, is translated by a sequence of application of rule $StructItem$.
The nested structures will be designed in the next project phase.

\subsection{Arrays}

The arrays store a fixed-size sequential collection of elements with the same type. 
They consist of contiguous memory locations, where the lowest address (0) 
corresponds to the first element and the highest address to the last element.
In B, a concrete array is a total function where the domain indexes the range 
elements. Declaration examples of arrays are:
\begin{itemize}
	\item $array1 \in (0..10) \to INT $	(a simple array)
	\item $array2 \in (1..10) \times (1..20) \times (1..30) \to INT $   (a multidimensional array)
	\item $\{1\mapsto1\mapsto4, 1\mapsto2\mapsto5\}  \in (INT \times INT) \times INT$ (an explicit array)
\end{itemize}

@@

This section shows the relation between arrays B0 and LLVM. The sequence of
elements are defining an array, getting an element and setting an
element.

@@

The following code declares a multidimensional global array called ``Array2'' 
with 6000 (10*20*30) integer numbers with 32 bits and initialized with zero.

\begin{llvmcode}
@Array2 = internal global [10 x [20 x [30 x i32]]] zeroinitializer, align 16
\end{llvmcode}

\VGM{It is not implemented yet, because it requires generating an XML example.}
The following example initialises an new array in C:

\begin{ccode}
int32_t arr[1][2][1]= {{{4},{5}}};
\end{ccode}

The syntax rule\footnote{The underlined element type can be expanded in others types definitions.} used to initialize an array in LLVM is:\\
$@<var\_name> = constant [<size> x \underline{<type>}] [<type> <value>]$.\\
The translation of the last example is:

\begin{llvmcode}
@arr=constant[1 x [2 x [1 x i32]]] [[2 x [1 x i32]] [[1 x i32] [i32 4],[1 x i32] [i32 5]]]
\end{llvmcode}

\subsubsection{Getting an element from array}

@@

\begin{llvmcode}
%7 = getelementptr [10 x [20 x [30 x i32]]]* @Array2, i32 0, i64 2, i64 3, i64 4
\end{llvmcode}

The first parameter defines the type and name of the variable, and the others are
indexes to select the element in the data type. The second parameter (i32 0)
selects the first element in the pointer (* @Array2). The last elements 
(i64 2, i64 3, i64 4) select the value from position (2,3,4) storing it in \%7.
This array is multidimensional, so several indexes must be declared separated
by comma to get the specific element.

@@

\begin{pascalcode} 
VAR xx, yy, zz IN
	xx=2;
	yy=3;
	zz=4;
	Array2(x,y,z):=7
END
\end{pascalcode}

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 2, 3, 4 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.  
\begin{figure}[h]
\begin{llvmcode}
%xx = alloca i32
%yy = alloca i32
%zz = alloca i32
store i32 2, i32* %xx
store i32 3, i32* %yy
store i32 4, i32* %zz
%1 = load i32* %zz
%2 = sext i32 %1 to i64
%3 = load i32* %yy
%4 = sext i32 %3 to i64
%5 = load i32* %xx
%6 = sext i32 %5 to i64
%7 = getelementptr[10x[20x[30xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7         ;setting an element
\end{llvmcode}
\end{figure}

\subsection{Rules for support Array}

@@

>>>>

\subsection{Translation of structures and arrays}
\label{sec:StructArray}

Two approaches are possible to support arrays and structures. An approach is 
importing reusable libraries of AtelierB that are reported in the official 
document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. 
This approach is supported, but the generated code is more complex and it needs 
more intermediate steps to manipulate the data of imported components. 
Moreover, this approach allows dynamic allocation for arrays, but it is 
strongly not recommended for safety-critical systems. A second approach enjoys 
the direct correspondence between B0 native types and LLVM and this approach 
is presented here. This section describes the translation of derived data 
types showed in section~\ref{sec:data}.

\subsubsection{Structure} 
The structure declaration defines a list of variables to be placed under one 
name, allowing several variables to be accessed by one base name. This example  
initializes a structure, and gets and sets one value of xx'balance.

\begin{pascalcode}
xx := rec(id:1, balance:10);
xx'balance := xx'balance -1;
\end{pascalcode}

The lines 1 and 2 are represented in translation in LLVM code:

@@

The $RecField$ rule gets an element value from the record. The same rule is 
used on the left side of attribution as a $LValue$ and right side of 
attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get 
respectively the position and type of $n$ element in the structure.

\begin{align*}
\begin{split}
  \trad{n}{\B{RecField}} \isdef \\
   & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
   &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   &\quad  \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
   &\quad  \sep  \ell  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

The attribution from record structure to a record structure variable, line 1 
from last B code example, is translated by a sequence of application of rule 
$RecField$. The nested structures will be designed in the next project phase.
%TODO after test the first examples.

\subsubsection{Arrays}

The arrays store a fixed-size sequential collection of elements with the same type. 
They consist of contiguous memory locations, where the lowest address (0) 
corresponds to the first element and the highest address to the last element.
In B, a concrete array is a total function where the domain indexes the range 
elements.

%TODO Attributions and initializations

@@

This section shows the relation between arrays B0 and LLVM. The sequence of
elements are defining an array, getting an element and setting an
element. The white code block and gray code block are respectively B code and LLVM code.

@@

The following code declares a multidimensional global array called ``Array2'' 
with 16 (2*2*4) integer numbers with 32 bits.

\begin{llvmcode}
@array2 = internal global [2 x [2 x [4 x i32]]]
\end{llvmcode}

\subsubsection{Initialing an array}
The following two attribution commands have the same semantic effect.

\begin{pascalcode}
array2 := ( 0..1 * 0..1 * 0..3 * {5});

array2 := { 0|->0|->0|->5 , 0|->0|->1|->5 , 
            0|->1|->0|->5 , 0|->1|->1|->5 , 
            1|->0|->0|->5 , 1|->0|->1|->5 , 
            1|->1|->0|->5 , 1|->1|->1|->5 ,
            0|->0|->2|->5 , 0|->0|->3|->5 ,
            0|->1|->2|->5 , 0|->1|->3|->5 , 
            1|->0|->2|->5 , 1|->0|->3|->5 , 
            1|->1|->2|->5 , 1|->1|->3|->5 };
\end{pascalcode}

An alternative way to represent these attributions cannot be using sequential attributions as follow:

\begin{pascalcode}
array2(0,0,0) := 5;
// A sequence of attributions ...
array2(1,1,3) := 5;
\end{pascalcode}

It is not possible, because B array attribution are atomic. The B array 
attribution are made in two steps. First, the new value is calculated and
copied to a temporal place. Second, the sequence of new values are setted 
in theirs address. Finally, considering these steps and the execution of
following code then the value of ``array2(0,0,1)'' is ``2''.

\begin{pascalcode}
 array2 := ( (0..1) * (0..1) * (0..3) * {1});
 array2 := { 0|->0|->0|->2 , 0|->0|->1|->(a3d(0,0,0)+1), ...  };
\end{pascalcode}

@@

\begin{llvmcode}
%7 = getelementptr [2 x [2 x [4 x i32]]]* @Array2, i32 0, i64 1, i64 1, i64 3
\end{llvmcode}

The first two parameters define the type and name of the variable, and the
others are indexes to select the element in the data type. The third parameter (i32 0)
selects the first element in the pointer (* @array2). The last elements 
(i64 1, i64 1, i64 3) select the value from position (1,1,3) storing it in \%7.
This array is multidimensional, so several indexes must be declared separated
by comma to get the specific element.

@@

\begin{pascalcode} 
VAR xx, yy, zz IN
	xx=0;
	yy=1;
	zz=3;
	array2(x,y,z):=7
END
\end{pascalcode}

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 2, 3, 4 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.  
\begin{figure}[h]
\begin{llvmcode}
%xx = alloca i32
%yy = alloca i32
%zz = alloca i32
store i32 0, i32* %xx
store i32 1, i32* %yy
store i32 3, i32* %zz
%1 = load i32* %zz
%2 = sext i32 %1 to i64
%3 = load i32* %yy
%4 = sext i32 %3 to i64
%5 = load i32* %xx
%6 = sext i32 %5 to i64
%7 = getelementptr[2x[2x[4xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7         ;setting an element
\end{llvmcode}
\end{figure}

\subsubsection{Rules for support array}

@@

%TODO Describe here the restrictions of declarations arrays
\paragraph{Relevant restrictions of work:}
\begin{itemize}
\item  B0 definition - An array must be initialized declaring all values 
to avoid no determinism.
\item B0 definition from CommenC - The operator overwrite from arrays is not supported, for example:

\begin{pascalcode}
array := array <+ {1|->2|->0|->4,  1|->2|->1|->5}
\end{pascalcode}

\end{itemize}

### 2014-04-25 DOCUMENTATION/specification.tex
rewriting the initialisation

\date{March 2014}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
\newminted{c}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}

@@

\item[arrays] In LLVM, an array type is defined by its size and the type of its
  elements, the main issues in encoding a B array type into LLVM being to
  determine the size of the array and multi-dimensionality.
Declaring an array type:
\[
\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..99) \times (0..99)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
\]
Initializing an array:
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{1 \mapsto  100,2 \mapsto 101}_{seq\ Map}\}}_{Array}
\]
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..100 \times 0..100  \times}_{seq\ Exp} \underbrace{\{1\}}_{Exp} }_{Product}\}}_{Array}
\]

Updating a selected element from array by a value of variable:
\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
\]

Setting a selected element from array to a variable:
\[ 
\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \]
\end{description}

>>>>

\date{March 2014}
\definecolor{bg}{rgb}{0.93,0.93,0.93}
\definecolor{white}{rgb}{1,1,1}
\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=white}

@@

\item[arrays] In LLVM, an array type is defined by its size and the type of its
  elements, the main issues in encoding a B array type into LLVM being to
  determine the size of the array and multi-dimensionality.
Declaring an array type:
\[
\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..99) \times (0..99)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
\]

There are three attributions cases related to array.
\subitem\textbf{Case 1.1} - Initializing an unique value to an array:
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2  }_{seq\ Exp} \times \underbrace{\{1\}}_{Exp} }_{Product}\}}_{Array}
\]
\subitem\textbf{Case 1.2} - Initializing different values to an array:
\[
\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(0 \mapsto 0) \mapsto  100,(0 \mapsto 1) \mapsto  101}_{seq\ Map}\}}_{Array}
\]

\subitem\textbf{Case 2.1} - Updating a selected element from array by a value of variable:
\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
\]

\subitem\textbf{Case 2.2} - Setting a selected element from array to a variable:
\[ 
\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem}
\]

\subitem\textbf{Case 3} - Attribution between arrays:
\[
\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
\]
\end{description}
